plyranges: a grammar for transforming genomics data

### 2018-05-14 paper.Rmd
tidying up - almost there

First, we gather the BigWig file and extract its annotation information and 
filter it to chromosome 10.

[ML: please explain why we are doing this. Something like: we need to
     get the range for chromosome 10, so that we can use it as a
     filter when reading. The first time I read this, I was not sure what
     'annotation information' and 'gather' meant here. Why say
     "annotation information" when the API uses "genome information"?]

@@

Then we read the BigWig file only extracting scores if they overlap chromosome
10. The annotation information from the file is automatically included (in this
case the hg19 genome build).

[ML: getting the genome info does not look automatic, because we are calling
	`set_genome_info()`; is that supposed to be here?]

@@

The `reduce_ranges()` operation is used to find coverage peaks 
across chromosome 10. We can manually set a threshold to restrict genomic
regions to have a coverage score of greater than 8, and
then merge nearby regions. The maximum coverage is computed over all the
coverage scores in the regions that were reduced.

[ML: I think the above text could do a better job of describing what
	`reduce_ranges()` does generally. You could say something like, after
	filtering for regions with a score greater than 8, we can reduce the
	individual runs to ranges representing the islands of coverage.]

@@

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved
via the `group_by_overlaps` operator.  We bin and count reads from a
H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
Consortium. The bins have a fixed width of 10000bp.

[ML: why not get the seqinfo from the BAM itself?]

We reuse the genome information from our BigWig file, as the
BAM file was aligned to the hg19 genome build. Alignments over
the mitochondrial genome are dropped using the `dropSeqlevels` function
from the Bioconductor package `GenomeInfoDb` @R-gdb.
[ML: seems like we should have a `select_seqlevels()` or something, so
     that we could do `select_seqlevels(-chrM)`.. somehow we have to
     explain to the reader why we are doing all of this seqlevel munging.]
```{r locations, eval = FALSE}
locations <- bw_file %>%
  get_genome_info() %>% 
  dropSeqlevels("chrM", pruning.mode = "coarse") %>%
  set_genome_info(genome = "hg19") 
```

To generate bins of width 10000bp we use the `tile` function from 
`GenomicRanges`:

```{r, eval = FALSE}
bins <- tile(locations, width = 10000)
```

Next we only read in alignments if they have a map quality score
greater than 20.
```{r, eval = FALSE}
alignments <- read_bam(h1_bam_sorted) %>%
    filter(mapq > 20) 
```

[ML: seems like a great opportunity to point out the deferred reading]

@@

[ML: there was a typo in the code: have you actually run this?]

```{r, eval = FALSE}
alignments_summary <- bins %>%
  group_by_overlaps(alignments) %>%
  summarize(n_reads = n())
```

@@

[ ML: there should be some quick summary paragraph here ]

>>>>

First, we extract the genome information from the BigWig file and filter
to get the range for chromosome 10.

First, we gather the BigWig file and extract its annotation information and 
filter it to chromosome 10. This range will be used as  a filter when reading
the file.

@@

Then we read the BigWig file only extracting scores if they overlap chromosome
10. We also add the genome build information to the resulting ranges.

@@

After filtering for regions with a coverage score greater than 8, we can
reduce individual runs to ranges representing the islands of coverage,
using the `reduce_ranges()` function. This function allows a summary
to be computer over each island: in this case we take the maximum of 
the scores to find the coverage peaks over chromosome 10.

@@

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved
via the `group_by_overlaps()` operator.  We bin and count reads from a
H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
Consortium. The bins have a fixed width of 10000bp.

We can directly obtain the genome information from the header of the
BAM file: in this case the reads were aligned to the hg19 genome build
and there are no reads overlapping the mitochondrial genome. To generate 
bins of width 10000bp we apply `tile` function from `GenomicRanges`
to the genomic coordinates:
```{r bins, eval = TRUE}
locations <- h1_bam_sorted %>%
  read_bam() %>% 
  get_genome_info()

bins <- locations %>% 
  tile(width = 10000L)
```

In this case using `read_bam()` in conjunction with `get_genome_info()`
only loads the genome information and nothing else.

Next we only read in alginments that overlap the genomic locations
we are interested in. The reading is deferred until after the filter is 
constructed: that is only alignments that pass the filter are loaded into memory.

```{r, eval = TRUE, cache = TRUE}
alignments <- h1_bam_sorted %>% 
    read_bam() %>%
    select(qwidth) %>%

    filter_by_overlaps(locations) %>%
```

@@

```{r, eval = FALSE}
alignments_summary <- bins %>%
  join_ %>%
  summarize(n_reads = n())
```

@@

We have shown how to create expressive and 
reproducible genomic workflows using the `plyranges` DSL. By
realising that the GRanges data model is tidy we have highlighted how to 
implement a grammar for performing genomic arithemtic, aggregation, restriction 
and merging. Our examples show that `plyranges` code is succinct and human
readable.
